---
title: "Getting Started with maplibr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with maplibr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`maplibr` is an R interface to `maplib`, a Rust RDF knowledge-graph engine
(the same engine the Python package `py_maplib` wraps). This vignette walks
through what's implemented today: building a graph, querying and updating it
with SPARQL, mapping data into RDF (both with OTTR templates and directly),
RDFS inference, and managing named graphs.

```{r setup}
library(maplibr)
```

## Creating a Model and reading triples

`Model` is the main entry point -- an in-memory RDF knowledge graph.

```{r create-and-read}
m <- Model$new()

m$reads('
  <http://example.org/alice> <http://www.w3.org/1999/02/22-rdf-syntax-ns#type> <http://example.org/Person> .
  <http://example.org/Person> <http://www.w3.org/2000/01/rdf-schema#subClassOf> <http://example.org/Agent> .
  <http://example.org/alice> <http://example.org/name> "Alice" .
', format = "ntriples")

m$size()
```

## Serializing back out

```{r writes}
cat(m$writes())
```

Turtle works too:

```{r writes-turtle}
cat(m$writes(format = "turtle"))
```

`$read()`/`$write()` are the file-based counterparts to
`$reads()`/`$writes()` -- `$read()` guesses the format from the file
extension (`.ttl`/`.nt`/`.xml`/`.rdf`) when `format` isn't given:

```{r read-write-file}
path <- tempfile(fileext = ".ttl")
m$write(path, format = "turtle")

m_from_file <- Model$new()
m_from_file$read(path)
m_from_file$size()
unlink(path)
```

## Querying with SPARQL

`$query()` runs a SPARQL query and returns the result as a data.frame.

```{r query-select}
m$query("SELECT ?s ?o WHERE { ?s <http://example.org/name> ?o }")
```

CONSTRUCT queries are supported too, returned as a combined
subject/predicate/object data.frame:

```{r query-construct}
m$query("
  CONSTRUCT { ?s <http://example.org/hasType> ?type }
  WHERE { ?s a ?type }
")
```

`bindings` pre-binds SPARQL variables to fixed `IRI`/`Literal` values before
the query runs -- useful inside a `FILTER`, though not as a `SELECT`-ed
column itself:

```{r query-bindings}
m$query(
  'SELECT * WHERE { VALUES ?name { "Alice" "Bob" } FILTER(?name = ?who) }',
  bindings = list(who = Literal("Bob"))
)
```

## Updating data with SPARQL UPDATE and INSERT

`$update()` runs a SPARQL UPDATE (a `DELETE`/`INSERT` ... `WHERE` modify
statement) against the Model in place. The bare `INSERT DATA { ... }` form
isn't supported yet by the underlying engine (see the table at the end) --
use an empty `WHERE {}` instead:

```{r update}
m$update('
  PREFIX ex: <http://example.org/>
  INSERT { ex:alice ex:greeting "Hello" } WHERE {}
')
m$query("SELECT ?o WHERE { <http://example.org/alice> <http://example.org/greeting> ?o }")
```

`$insert()` runs a SPARQL `CONSTRUCT` query and adds its results as new
triples, optionally into a different graph than it queried:

```{r insert}
m$insert(
  "CONSTRUCT { ?s <http://example.org/hasType> ?type } WHERE { ?s a ?type }",
  target_graph = "urn:graph:types"
)
m$size("urn:graph:types")
```

## Mapping tabular data into RDF with OTTR templates

`$map()` expands an OTTR template against a data.frame, adding the resulting
triples to the Model. Templates can be built programmatically...

```{r map-programmatic}
ex <- Prefix("http://example.org/", "ex")
my_value <- Variable("MyValue")
template <- Template(
  iri = suf(ex, "HasNickname"),
  parameters = list(Parameter(my_value)),
  instances = list(Triple(suf(ex, "alice"), suf(ex, "nickname"), my_value))
)

m$map(template, data.frame(MyValue = c("Ally", "Al")))
m$query("SELECT ?o WHERE { <http://example.org/alice> <http://example.org/nickname> ?o }")
```

...or from an stOTTR document string, matching `py_maplib`'s own syntax (see
the root project's README for the OTTR template language itself):

```{r map-doc-string}
doc <- '
@prefix ex:<http://example.org/>.
ex:HasFriend [?Friend] :: {
  ottr:Triple(ex:alice, ex:knows, ?Friend)
} .
'
m$add_template(doc)
m$map("http://example.org/HasFriend", data.frame(Friend = "http://example.org/bob"))
m$query("SELECT ?o WHERE { <http://example.org/alice> <http://example.org/knows> ?o }")
```

## Mapping data directly to triples, without a template

For data that doesn't need OTTR's templating power, several `$map_*()`
methods add triples directly instead. `$map_df()` turns a data.frame's
columns into predicates, one row per subject:

```{r map-df}
readings <- data.frame(city = c("Oslo", "Bergen"), temp_c = c(14, 16))
m_direct <- Model$new()
m_direct$map_df(readings)
m_direct$query("SELECT ?p ?o WHERE { ?s ?p ?o }")
```

`$map_triples()` maps subject/predicate/object columns directly (or just
subject/object, with a constant `predicate`):

```{r map-triples}
m_triples <- Model$new()
m_triples$map_triples(
  data.frame(subject = "http://example.org/alice", object = "http://example.org/bob"),
  predicate = "http://example.org/knows"
)
m_triples$query("SELECT ?s ?p ?o WHERE { ?s ?p ?o }")
```

`$map_json_string()`/`$map_xml_string()` (and their file-based
`$map_json()`/`$map_xml()` counterparts) apply a fixed convention
(Facade-X) to a whole JSON or XML document instead:

```{r map-json}
m_json <- Model$new()
m_json$map_json_string('{"name": "Alice", "age": 30}')
m_json$query("SELECT ?p ?o WHERE { ?s ?p ?o }")
```

## RDFS inference

`alice` is a `Person`, and `Person` is a subclass of `Agent` -- RDFS
inference should let us derive that `alice` is also an `Agent`.

```{r infer}
n <- m$infer_rdfs()
n
```

Inferred triples are *transient* -- kept separately from the main graph, and
only visible to `$query()` when asked for explicitly (`$writes()` doesn't
currently support this at all, a core-engine limitation, not an R-wrapper
one -- see the "Upstream issues" section of the package's development notes):

```{r infer-transient}
without <- m$query("SELECT * WHERE { ?s ?p ?o }")
with_transient <- m$query("SELECT * WHERE { ?s ?p ?o }", include_transient = TRUE)
nrow(without)
nrow(with_transient)
```

## Named graphs

Triples can be routed into named graphs, copied between Models, and split
back out.

```{r graphs}
m$reads(
  '<http://example.org/bob> <http://example.org/name> "Bob" .',
  format = "ntriples",
  graph = "urn:graph:extra"
)
m$size("urn:graph:extra")
m$size() # default graph, unaffected

m2 <- Model$new()
m2$add_graph(m)
m2$size()

sprout <- m$detach_graph(graph = "urn:graph:extra")
sprout$size()
m$size("urn:graph:extra") # emptied by the detach
```

## Prefixes and indexing

```{r prefixes-index}
m$add_prefixes(list(ex = "http://example.org/"))
m$create_index()
```

## Visualizing a graph

`maplibr` has no graph-visualization function of its own, but a query result
is a plain data.frame, so any R graph package works directly on it.

```{r graph-plot, fig.width=6, fig.height=4.5}
if (requireNamespace("igraph", quietly = TRUE)) {
  triples <- m$query("SELECT ?s ?p ?o WHERE { ?s ?p ?o }")
  short_label <- function(iri) sub(".*[/#]", "", iri)

  edges <- data.frame(from = short_label(triples$s), to = short_label(triples$o))
  g <- igraph::graph_from_data_frame(edges, directed = TRUE)
  igraph::E(g)$label <- short_label(triples$p)

  plot(
    g,
    edge.label = igraph::E(g)$label,
    vertex.size = 32,
    vertex.color = "lightblue",
    vertex.label.cex = 0.9,
    edge.arrow.size = 0.5,
    layout = igraph::layout_with_fr(g)
  )
}
```

## Pipe-friendly usage

Every `Model` method above is also available as a standalone `model_*()`
function taking the Model as its first argument, so `m |> model_query(...)`
works as an alternative to `m$query(...)` -- both are equivalent and stay
available side by side (the same convention `arrow` and `torch` use for
their own R6-based objects). This rewrites the "Creating a Model" example
from the top of this vignette as one pipeline:

```{r pipe-example}
m_piped <- Model$new() |>
  model_reads('
    <http://example.org/alice> <http://www.w3.org/1999/02/22-rdf-syntax-ns#type> <http://example.org/Person> .
    <http://example.org/alice> <http://example.org/name> "Alice" .
  ', format = "ntriples") |>
  model_update('
    PREFIX ex: <http://example.org/>
    INSERT { ex:alice ex:greeting "Hello" } WHERE {}
  ')

m_piped |> model_query("SELECT ?p ?o WHERE { <http://example.org/alice> ?p ?o }")
```

The `model_` prefix (rather than bare `read()`/`write()`/`map()`/`update()`)
is deliberate: those names would mask `base::write`, `purrr::map`, and the
`stats::update` S3 generic if this package were attached alongside a normal
tidyverse session.

## What doesn't work yet, and why

| Method(s) | Status |
|---|---|
| `$writes(include_transient = TRUE)` | Not supported for any format -- the underlying engine's triple-serialization code has no way to include transient/inferred triples, unlike its query path. A core-engine limitation, not specific to this package. |
| `$validate()` (SHACL) | Not implemented -- also blocked on a licensing question (the root project's README describes SHACL as proprietary, but its source is present and unrestricted in this checkout; not a decision this package can make unilaterally). |
| `$serialize()` / `$deserialize()` | Wired up correctly, but panic at runtime -- the on-disk storage engine they depend on is a closed-source stub in this checkout. |
| `$add_udf()` | Not implemented -- the underlying engine's UDF callback mechanism is CPython-specific end to end (it calls into Python's own `io`/`polars` modules to invoke the callback), with no way for a non-Python binding to register one instead. A core-engine limitation. |
| bare `INSERT DATA { ... }` / `DELETE DATA { ... }` (in `$update()`) | Panics instead of raising a normal error -- unimplemented in the underlying engine's SPARQL UPDATE handling. Use `INSERT { ... } WHERE {}` / `DELETE { ... } WHERE {}` instead, which works correctly. A core-engine limitation. |
| CIM XML, HDT, JSON-LD | Not ported. |

See the package's development notes for the full, current list of known
issues and their upstream/local status.
